---
title: "Supplementary Material: Evolutionary transition between bee- and hummingbird-pollination in *Salvia*: comparing means, variances and covariances of corolla traits"
author: "Santiago Benitez-Vieyra, César A. Domínguez, Jessica Pérez-Alquicira, Federico D. Sazatornil, Karina Boege, Rubén Pérez-Ishiwara & Juan Fornoni"

output: 
  pdf_document:
    fig_width: 7.5
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, fig.align = 'center', 
                      fig.pos = "p",out.extra = ' ', message=FALSE, fig.show="hold")
options(knitr.kable.NA = '')
```

This is a pdf version of an R Markdown, the original routines can be downloaded from https://github.com/santiagombv/salvia.matrix.evol, along with the raw data from 18 *Salvia* species.   

## 1. Data imput.
```{r}
# read data
sal <- read.csv("flower_measures.csv", header = TRUE)
sal$plant <- as.factor(sal$plant)
sal$flower <- as.factor(sal$flower)

# log transformation
sal[, paste("log", names(sal[, 5:10]), 
            sep = ".")] <- apply(sal[, 5:10], 2, FUN = log)

# list of data frames
sal.prev <- split(sal, paste(sal$sp, sal$pop, sep = "_"))

# Individual means
# obtain a single data frame (sal.base)
# and a list of data frames (sal.list)
resum<-function(x){
  if(is.factor(x)==TRUE) y <- unique(droplevels(x)) else y <- mean(x, na.rm=T)
  return(y)
}

sal.list <- list(21)
for(i in 1:21) sal.list[[i]] <- aggregate(sal.prev[[i]][, -4], 
                                           by = list(sal.prev[[i]][, "plant"]), 
                                           FUN=resum)[, -c(1)]
sal.base <- do.call("rbind", sal.list) 
rm(sal, sal.prev)
```

\pagebreak

## 2. Mean trait space (morphospace).
### 2.1. Space building.
   
```{r}
library(treebase)
library(phytools)
library(knitr)
library(vegan)

# obtain species means
pop.means <- aggregate(sal.base[, -3], by = list(sal.base$pop), FUN = resum)[, -1]
sp.means <- aggregate(pop.means[, -2], by = list(pop.means$sp), FUN = resum)[, -1]

# obtain phylogenetic tree from Treebase and pruning
T1 <- search_treebase("S15364", by="id.study")[[1]]

salvia.species <- c("Salvia_atrocyanea", "Salvia_calolophos", "Salvia_cinnabarina",
                    "Salvia_cuspidata", "Salvia_elegans", "Salvia_fulgens",
                    "Salvia_guaranitica", "Salvia_iodantha", "Salvia_lavanduloides",
                    "Salvia_longispicata", "Salvia_mexicana", "Salvia_misella",
                    "Salvia_pallida", "Salvia_personata", "Salvia_polystachya",
                    "Salvia_purpurea", "Salvia_stachydifolia", "Salvia_thyrsiflora")
Tsub1 <- drop.tip(T1, setdiff(T1$tip.label, salvia.species))
Tsub1$tip.label <- c("calo", "cusp", "fulg", "long", "lava", "guar", 
                     "pall", "atro", "mexi", "poly", "ioda", "purp", 
                     "stac", "pers", "thyr", "cinn", "eleg", "mise")

# phylogenetic principal components
rownames(sp.means) <- sp.means$sp
phyPC <- phyl.pca(tree=Tsub1, Y = sp.means[, 8:13])
result_pc <- as.data.frame(rbind(diag(phyPC$Eval), 
                                 diag(phyPC$Eval)/sum(diag(phyPC$Eval))*100,
                                 cumsum(diag(phyPC$Eval)/sum(diag(phyPC$Eval))*100)))
row.names(result_pc) <- c("eigenvalues", "eigenvalues %", "cummulative %")
kable(result_pc, digits = 3)

# vector of pollination syndromes (m = bee-pollinated; o = bird-pollinated)
synd <- c("o", "m", "o", "m", "o", "o", "o", "o", "m", 
          "m", "o", "m", "m", "m", "m", "o", "m", "m")

# mean trait space (morphospace). Plot
colores <- synd
colores[colores == "o"] <- "red"; colores[colores == "m"] <- "blue"
plot(phyPC$S[, 1]*(-1), phyPC$S[, 2]*(-1), pch = 19, col = colores,
     xlab = "principal component 1", ylab = "principal component 2")
ordihull(phyPC$S[, c(1,2)]*(-1), groups = synd, draw = "polygon", 
         alpha = 50, col = c("blue", "red"))

```



**Figure 1.** Mean trait space (morphospace) of *Salvia* corolla traits, showing the first two phylogenetic principal components. Red dots indicate hummingbird-pollinated *Salvia* species, while blue dots correspond to bee-pollinated species. The same colors indicate the convex hulls.   

\pagebreak

### 2.2. Diferences in corolla morphology between pollination syndromes.

```{r}
library(nlme)
library(vegan)
library(geiger)

# Phylogenetic generalized least squares (PGLS)
dataphyPC <- cbind.data.frame(phyPC$S, synd) 
s1 <- gls(PC1 ~ synd, data = dataphyPC, correlation = corPagel(1, Tsub1)) #PC1
kable(summary(s1)$tTable, digits = 3) # only coefficients

s2 <- gls(PC2 ~ synd, data = dataphyPC, correlation = corPagel(1, Tsub1)) #PC2
kable(summary(s2)$tTable, digits = 3) # only coefficients

# PERMANOVA
eu.distances <- dist(phyPC$S)
permanova.morpho <- adonis2(eu.distances ~ synd, permutations = 9999)
kable(permanova.morpho, digits = 3)

# test for diferences in dispersion between groups
disp.morpho <- betadisper(eu.distances, group=synd)
res.morpho <- permutest(disp.morpho, permutations = 9999)
kable(res.morpho$tab, digits = 3)

# Phylogenetic MANOVA (geiger)
sixtraits <- sp.means[, 8:13]
synd2 <- as.factor(synd)
names(synd2) <-rownames(sp.means)
fit1 <- aov.phylo(sixtraits ~ synd2, phy = Tsub1, nsim = 1000)

```

\pagebreak

## 3. Covariance space.
### 3.1. Space building.   

```{r}
library(evolqg)
library(bios2mds)
library(geometry)
library(plot3D)
library(knitr)

# list of P matrices per population
pmat <- list(21) 
for(i in 1:21) pmat[[i]] <- cov(sal.list[[i]][, 10:15], use= "pairwise.complete.obs")

# list of P matrices per species
N <- numeric(21); for(i in 1:21) N[i] <- nrow(sal.list[[i]])
salvia_p <- list(atro = pmat[[1]], calo = pmat[[2]], cinn = pmat[[3]],
                 cusp = (pmat[[4]]*N[4]+pmat[[5]]*N[5])/(N[4]+N[5]), 
                 eleg = (pmat[[6]]*N[6]+pmat[[7]]*N[7])/(N[6]+N[7]), 
                 fulg = (pmat[[8]]*N[8]+pmat[[9]]*N[9])/(N[8]+N[9]), 
                 guar = pmat[[10]], ioda = pmat[[11]], lava = pmat[[12]],
                 long = pmat[[13]], mexi = pmat[[14]], mise = pmat[[15]],
                 pall = pmat[[16]], pers = pmat[[17]], poly = pmat[[18]],
                 purp = pmat[[19]], stac = pmat[[20]], thyr = pmat[[21]])
rm(pmat)

# obtain Riemannian distances
rdist <- as.dist(MatrixDistance(cov.x = salvia_p, distance = "RiemannDist"))

# perform Principal Coordinate Analysis (PCoA)
biommds <- mmds(active = as.matrix(rdist), pc = 17)

# eigenvalues and relative eigenvalues from PCoA
result_pcoa <- rbind(biommds$eigen[1:7], 
                     biommds$eigen.perc[1:7],
                    cumsum(biommds$eigen.perc[1:7]))
colnames(result_pcoa) <- paste("PCo", c(1:7), sep = "")
row.names(result_pcoa) <- c("eigenvalues", "eigenvalues %", "cummulative %")
kable(result_pcoa, digits = 3)

# vector of pollination syndromes (m = bee-pollinated; o = bird-pollinated)
synd <- c("o", "m", "o", "m", "o", "o", "o", "o", "m", 
          "m", "o", "m", "m", "m", "m", "o", "m", "m")

# convex hulls 
ps1 <- as.matrix(biommds$coord[synd == "o", c(1,2,3)])
ts1 <- t(convhulln(ps1)) 
ps2 <- as.matrix(biommds$coord[synd == "m", c(1,2,3)])
ts2 <- t(convhulln(ps2)) 

# Covariance space. 3D plot
colores <- character(18)
colores[synd=="o"] <- "red4"; colores[synd=="m"] <- "blue4"
scatter3D(biommds$coord[,1], biommds$coord[,2], biommds$coord[,3],
          colvar = NULL, theta = 135, phi = 30, bty="g", cex.lab = 0.8,
          col = colores, ticktype = "detailed", pch = 16, r=50,
          xlab = "principal coordinate 1",
          ylab = "principal coordinate 2",
          zlab = "principal coordinate 3")
for(i in 1:ncol(ts1)){
  polygon3D(ps1[ts1[,i],1],ps1[ts1[,i],2],ps1[ts1[,i],3],col=rgb(1,0,0,0.05),
            facets=T, border = "grey", add=T) 
}
for(i in 1:ncol(ts2)){
  polygon3D(ps2[ts2[,i],1],ps2[ts2[,i],2],ps2[ts2[,i],3],col=rgb(0,0,1,0.05),
            facets=T, border = "grey", add=T) 
}

```



**Figure 2.** **P**-matrix space of *Salvia* corolla traits, showing the first three principal coordinates. Red dots indicate hummingbird-pollinated *Salvia* species, while blue dots correspond to bee-pollinated species. The same colors indicate the convex hulls.   


\pagebreak


### 3.2. Correlation between covariance space axes and original variances and covariances.   

```{r}
library(ggcorrplot)

# data frame of covariances
Pdat <- matrix(rep(0,378), 18, 21)
for(i in 1:18) Pdat[i, ] <- salvia_p[[i]][lower.tri(salvia_p[[i]], diag = T)]
colnames(Pdat) <- c("lll.lll", "lll.llw", "lll.ull", "lll.ulw", "lll.ctl", "lll.ctw",
                    "llw.llw", "llw.ull", "llw.ulw", "llw.ctl", "llw.ctw", "ull.ull", 
                    "ull.ulw", "ull.ctl", "ull.ctw", "ulw.ulw", "ulw.ctl", "ulw.ctw",
                    "ctl.ctl", "ctl.ctw", "ctw.ctw")
Pdat <- as.data.frame(Pdat)

# merge data frames
mix.dat <- data.frame(biommds$coord[, c(1:7)], Pdat)

# correlations
kable(t(cor(mix.dat)[1:7, names(Pdat)]), digits = 3)

# plot
o <- c(20,18,17,15,14,13,11,10,9,8,6,5,4,3,2,21,19,16,12,7,1)
ggcorrplot(cor(mix.dat)[1:3, names(Pdat)[o]], lab = FALSE, lab_size = 2,
           colors = c("#5e3c99", "white", "#e66101"))

```



**Figure 3.** Pearson correlations between the principal coordinates and the variances and covariances of the **P**-matrices.   

\pagebreak

### 3.3. Divergence in the covariance space between pollination syndromes.  

```{r}
# PGLS
dataphyPCoA <- cbind.data.frame(biommds$coord, synd)
str.cor <- corPagel(1, Tsub1)

t1 <- gls(PC1 ~ synd, data = dataphyPCoA, correlation = str.cor, 
          control= list(opt="optim")) #PCoA1
kable(summary(t1)$tTable, digits = 3) # only coefficients

t2 <- gls(PC2 ~ synd, data = dataphyPCoA, correlation = str.cor,
          control= list(opt="optim")) #PCoA2
kable(summary(t2)$tTable, digits = 3) 

t3 <- gls(PC3 ~ synd, data = dataphyPCoA, correlation = str.cor,
          control= list(opt="optim")) #PCoA3
kable(summary(t3)$tTable, digits = 3) 

t4 <- gls(PC4 ~ synd, data = dataphyPCoA, correlation = str.cor,
          control= list(opt="optim"))#PCoA4
kable(summary(t4)$tTable, digits = 3) 

t5 <- gls(PC5 ~ synd, data = dataphyPCoA, correlation = str.cor,
          control= list(opt="optim"))#PCoA5
kable(summary(t5)$tTable, digits = 3) 

t6 <- gls(PC6 ~ synd, data = dataphyPCoA, correlation = str.cor,
          control= list(opt="optim"))#PCoA6
kable(summary(t6)$tTable, digits = 3) 

t7 <- gls(PC7 ~ synd, data = dataphyPCoA, correlation = str.cor,
          control= list(opt="optim"))#PCoA7
kable(summary(t7)$tTable, digits = 3) 

# PERMANOVA
permanova.cov <- adonis2(rdist ~ synd, permutations = 9999)
kable(permanova.cov, digits = 3)

# test for diferences in dispersion between groups
disp.cov <- betadisper(rdist, group=synd)
res.cov <- permutest(disp.cov, permutations = 9999)
kable(res.cov$tab, digits = 3)

# Phylogenetic MANOVA (geiger)
sevenaxes <- dataphyPCoA[, 1:7]
fit2 <- aov.phylo(sevenaxes ~ synd2, phy = Tsub1, nsim = 1000)

```
   
   
   
\pagebreak   

## 4. Multivariate Phylogenetic Signal.   

```{r}
library(phylocurve)

# multivariate phylogenetic signal in six corolla traits
names(sp.means)[1] <- "species"
null1 <- evo.model(tree = Tsub1, Y = sp.means[, c(1, 8:13)], method = "Pairwise ML")
K.mult(null1, plot = FALSE)

# multivariate phylogenetic signal in the first 
# seven axes of the coavariance space
pcoa_phy <- biommds$coord[, 1:7]
pcoa_phy$species <- rownames(biommds$coord)
null2 <- evo.model(tree = Tsub1, Y = pcoa_phy, method = "Pairwise ML")
K.mult(null2, plot = FALSE)

```



\pagebreak


## 5. Comparison of morphological and variance-covariance spaces.   

```{r}

# Distance matrices
dist.coph <- cophenetic.phylo(Tsub1)
dist.cova <- as.matrix(rdist)
dist.morp <- as.matrix(dist(phyPC$S[, 1:6], upper = T, diag = T))

# partial Mantel function
partial.cor <- function(xdis, ydis, zdis){
  xdis <- as.dist(xdis)
  ydis <- as.vector(as.dist(ydis))
  zdis <- as.vector(as.dist(zdis))
  rxy <- cor(as.vector(xdis), ydis)
  rxz <- cor(as.vector(xdis), zdis)
  ryz <- cor(ydis, zdis)
  part.cor <- (rxy - rxz * ryz)/sqrt(1 - rxz * rxz)/sqrt(1 - ryz * ryz)
  return(part.cor)
}

# phylogenetic permutations
phyloProb<-function(phy, k=2) {
  pd<-cophenetic.phylo(phy)
  pdr<-pd/max(pd)
  s<-k-pdr
  p<-s/rowSums(s)
  p
}

# Permutes species according to phylogentic tree
# returns tip names in permuted order
# (from Harmon and Glor, 2010)
phyloPermute<-function(phy, k=1) {
  p<-phyloProb(phy, k)
  tt<-rownames(p)
  nsp<-length(tt)
  order<-sample(1:nsp, replace=F)
  ttnew<-character(nsp)
  cpm<-p
  for(j in order[-nsp]) {
    cpm<-cpm/rowSums(cpm)
    rr<-which(rownames(cpm)==tt[j])
    pp<-cpm[rr,]
    s2<-sample(names(pp), size=1, replace=T, prob=pp)
    slot<-which(tt==s2)
    rc<-which(colnames(cpm)==s2)
    ttnew[slot]<-tt[j]
    cpm<-cpm[-rr,-rc]
  }
  ttnew[which(ttnew=="")]<-tt[order[nsp]]
  ttnew
}	

# observed value
obs.value <- partial.cor(xdis = dist.cova, ydis = dist.morp, zdis = dist.coph)

# pseudo-values
partial.mantel <- replicate(1000, partial.cor(xdis = dist.cova[phyloPermute(phy=Tsub1),
                                                              phyloPermute(phy=Tsub1)],
                                               ydis = dist.morp, zdis = dist.coph))
# significance
p.value <- length(partial.mantel[partial.mantel>obs.value])/1000
kable(data.frame(r = obs.value, P = p.value), digits = 3)

```



\pagebreak

## 6. Numerical simulations   

### 6.1. Projecting proportional matrices into the original P-matrix space.
The function *prop_matrix* performs the eigendecomposition of a covariance matrix, multiplies the eigenvalues by a given value (proportionality constant) and obtains a new covariance matrix with the new eigenvalues and the same eigenvectors than the original one.   

```{r}

# prop_matrix function arguments:
# mat   A covariance matrix
# pc   Proportionality constant
# Output: a new matrix

prop_matrix <- function(mat, pc){
  ei <- eigen(mat)
  newmat <- ei$vectors%*%diag(ei$values*pc)%*%solve(ei$vectors)
  return(newmat)
}

# Setting a vector of proportionality constants
prop.values <- c(rep(0.5, 18), rep(2, 18))

# Creating new matrices from the original P-matrices.
matrix.index <- rep(names(salvia_p), 2)
Z <- list(18*2)
for(i in 1:36){
  Z[[i]] <- prop_matrix(salvia_p[[matrix.index[i]]], pc = prop.values[i])
  names(Z)[i] <- paste(matrix.index[i], prop.values[i], sep = "_")
}

# Merge original and new matrices
Pmat.add1 <- c(salvia_p, Z) 

# Obtain Riemannian distances
distances1 <- as.dist(MatrixDistance(cov.x = Pmat.add1, distance = "RiemannDist"))

# Projecting the new matrices into the original P-matrix space 
SUP1 <- as.matrix(distances1)[19:54,-c(19:54)]
pru1 <- mmds.project(mmds=biommds, sup = SUP1, pc = 17)

# 3D plot
scatter3D(biommds$coord[,1], biommds$coord[,2], biommds$coord[,3],
          colvar = NULL, theta = 45, phi = 30, col = colores, bty="g",
          ticktype = "detailed", pch = 16, r=50, cex = 1.2, cex.lab = 0.8,
          xlab = "principal coordinate 1",
          ylab = "principal coordinate 2",
          zlab = "principal coordinate 3",
          xlim = c(-8, 8), ylim = c(-8, 8), zlim = c(-8, 8))
arrplot <- function(x){
  arrows3D(pru1$coord[x,1], pru1$coord[x,2], pru1$coord[x,3], 
           pru1$coord[x+18,1], pru1$coord[x+18,2], pru1$coord[x+18,3], 
           add= T, type = "triangle", col = "black", lwd = 1.2) 
}
for(i in 1:18) arrplot(i)

```



**Figure 4.** Projection of simulated matrices into the **P**-matrix space. The start of the arrow corresponds to a matrix having half the total variance as the original matrix. The end of the arrow corresponds to a matrix having double the total variance as the original matrix.   

\pagebreak   


### 6.2. Projecting matrices with diferent integration into the original P-matrix space.  

The function *bs_matrix* performs the eigendecomposition of a covariance matrix. Then, it redistributes the variance among eigenvalues, assigning a given amount to the leading (first) eigenvalue and distributing the remaining variance among the rest. The total amount of variance  is not changed. Finally, it obtains a new covariance matrix with the new eigenvalues and the same eigenvectors than the original one.   
   
Because only the first eigenvalue is trasformed, the function simmulate a "broken stick" eigenvalue distribution. Thus, integration increases with the proportion of the variance explained by the leading eigenvalue. Maximum integration (measured as relative eigenvalue variance following Wagner 1984 and  Cheverud *et al.* 1989) is attained when the first eigenvalue accounts for all variance. Given that **P**-matrices include six traits, minimum integration is attained when the first eigenvalue explains 16.67% of the total variance. 
   
Integration is often estimated from correlation matrices. Thus, our results are not comparable with most studies, but we are confident about the rationale of using the amount of variance explained by the leading eigenvalue as an estimate of phenotypic integration.   
    
     
```{r}

# Proportion of the variance explained by the leading (first) eigenvalue
ei <- numeric(18)
for(i in 1:18) {
  Z <- eigen(salvia_p[[i]])$values
  ei[i] <- Z[1]/sum(Z)
}

# bs_matrix function arguments:
# mat   A covariance matrix
# prop.var  Proportion of the total variance explained by the leading (first) 
# eigenvalue
# Output: a new matrix

bs_matrix <- function(mat, prop.var){ 
  ei <- eigen(mat)
  SL <- sum(ei$values)
  dim <- nrow(mat)
  L1 <- SL*prop.var
  Lk <- ei$values[-1]/sum(ei$values[-1])*(SL-L1)
  d <- c(L1, Lk)
  newmat <- ei$vectors%*%diag(d)%*%solve(ei$vectors)
  return(newmat)
}

# vector of proportions of the variance explained by the first eigenvalue
leading.ei <- c(rep(0.25, 18), rep(0.85, 18))

# Creating new matrices from the original P-matrices.
matrix.index <- rep(names(salvia_p), 2)
Y <- list(18*2)
for(i in 1:36){
  Y[[i]] <- bs_matrix(salvia_p[[matrix.index[i]]], prop.var = leading.ei[i])
  names(Y)[i] <- paste(matrix.index[i], leading.ei[i], sep = "_")
}

# Merge into a list the original and new matrices
Pmat.add2 <- c(salvia_p, Y) 

# Obtain Riemannian distances
distances2 <- as.dist(MatrixDistance(cov.x = Pmat.add2, distance = "RiemannDist"))

# Projecting the new matrices into the original P-matrix space 
SUP2 <- as.matrix(distances2)[19:54,-c(19:54)]
pru2 <- mmds.project(mmds=biommds, sup = SUP2, pc = 17)

# 3D plot
scatter3D(biommds$coord[,1], biommds$coord[,2], biommds$coord[,3],
          colvar = NULL, theta = 45, phi = 30, col = colores, bty="g",
          ticktype = "detailed", pch = 16, r=50, cex = 1.2, cex.lab = 0.8,
          xlab = "principal coordinate 1",
          ylab = "principal coordinate 2",
          zlab = "principal coordinate 3",
          xlim = c(-8, 8), ylim = c(-8, 8), zlim = c(-8, 8))
segplot <- function(x){
  segments3D(pru2$coord[x,1], pru2$coord[x,2], pru2$coord[x,3], 
             biommds$coord[x,1], biommds$coord[x,2], biommds$coord[x,3],
             add= T, lwd=1.2) 
}
arrplot <- function(x){
  arrows3D(biommds$coord[x,1], biommds$coord[x,2], biommds$coord[x,3],
           pru2$coord[x+18,1], pru2$coord[x+18,2], pru2$coord[x+18,3], 
           add= T, lwd= 1.2, col = "black", type = "triangle") 
}

for(i in 1:18) segplot(i)
for(i in 1:18) arrplot(i)


```



**Figure 5.** Projection of simulated matrices into the **P**-matrix space. The start of the arrow corresponds to a matrix where the first eigenvector accounts for 25 percent of the total variance. The end of the arrow corresponds to a matrix where the first eigenvector accounts for 85 percent of the total variance.   

\pagebreak


### 6.3. Average vector correlations.   

Vector correlations (cosine of the angle between vectors) between the first seven coordinates (PCo) from Principal Coordinate Analysis and the direction in covariance space where the total variance increases or decreases (proportionality changes) and the direction where the magnitude of phenotypic integration increases or decreases (integration changes).   

```{r}
# angle.calc arguments:
# coord data frame of vector doordinates
# from rowname of the vector origin
# to rowname of the vector end
# dim number of dimensions

angle.calc <- function(from, to, coord, dim){
  vec <- as.numeric(coord[to, ] - coord[from, ])
  axes <- list(dim)
  for(i in 1:dim){
    axes[[i]] <- rep(0, dim)
    axes[[i]][i] <- 1
  }
  cosines <- numeric(dim)
  for(i in 1:dim){
    cosines[i] <- (t(vec)%*%axes[[i]])/sqrt((t(vec)%*%vec)*(t(axes[[i]])%*%axes[[i]]))
  }
  angles <- acos(cosines)
  res <- c(cosines, angles)
  names(res) <- c(paste("cos", c(1:dim), sep = "."),
                  paste("ang", c(1:dim), sep = "."))
  return(res)
}

# vector correlation. Proportionality simulations
ori1 <- paste(names(salvia_p), 0.5, sep = "_")
fin1 <- paste(names(salvia_p), 2, sep = "_")
ANG1 <- list(18)
for(i in 1:18) ANG1[[i]] <- angle.calc(from = ori1[i], to=fin1[i], coord = pru1$coord, dim = 17)

ANG.PROP <- do.call(rbind, ANG1)
vector.corr1 <- colMeans(ANG.PROP)[1:7]
names(vector.corr1) <- NULL

# vector correlation. Integration simulations
ori2 <- paste(names(salvia_p), 0.25, sep = "_")
fin2 <- paste(names(salvia_p), 0.85, sep = "_")
ANG2 <- list(18)
for(i in 1:18) ANG2[[i]] <- angle.calc(from = ori2[i], to=fin2[i], coord = pru2$coord, dim = 17)

ANG.INT <- do.call(rbind, ANG2)
vector.corr2 <- colMeans(ANG.INT)[1:7]
names(vector.corr2) <- NULL

kable(data.frame(PCo = paste("PCo", c(1:7), sep = "."),
                 prop.changes = vector.corr1,
                 integ.changes = vector.corr2), digits = 3)
```



\pagebreak


### 6.4. Check: Does the proportion of the total variance explained by the first eigenvalue account for phenotypic integration?   

```{r, fig.height = 5}
library(sciplot)

# proportion of the total variance explained by the first eigenvalue
LEint <- function(x){
  EI<- eigen(x, symmetric = TRUE)$values
  return(EI[1]/sum(EI))
}

# Hansen-Houle integration index 
HHint <- function(x){
  k <- nrow(x)
  a <- numeric(length=k)
  for(i in 1:k) a[i] <- (solve(x)[i,i]*x[i,i])^(-1)
  int <- 1 - a
  return(mean(int))
}

# Wagner-Cheverud integration index
WCint <- function(x){
  EI <- eigen(x, symmetric = TRUE)$values
  k <- ncol(x)
  INT <- ((sum((EI - mean(EI))^2)/k))/(k - 1)
  return(INT)
}

# base building after simulations
integ_HH <- numeric(54)
integ_WC <- numeric(54)
integ_LE <- numeric(54)
for(i in 1:54){
  integ_HH[i] <- HHint(x = Pmat.add2[[i]])
  integ_WC[i] <- WCint(x = cov2cor(Pmat.add2[[i]]))
  integ_LE[i] <- LEint(x = Pmat.add2[[i]])
} 
dat_integ <- data.frame(type = c(rep("original", 18), rep("0.25", 18), rep("0.85", 18)),
                        sp = rep(rownames(biommds$coord), 3),
                        integ_LE, integ_HH, integ_WC)

# plot
layout(matrix(1:2,1,2))
lineplot.CI(x.factor = type, response = integ_HH, group = sp, cex = 0.5,
            data = dat_integ, pch = rep(19, 18), lty = rep(1,18),
            legend = F, ylab = "Hansen-Houle integration index",
            xlab = "matrix type")
lineplot.CI(x.factor = type, response = integ_WC, group = sp, cex = 0.5,
            data = dat_integ, pch = rep(19, 18), lty = rep(1,18),
            legend = F, ylab = "Wagner-Cheverud integration index",
            xlab = "matrix type")
layout(1)

```



**Figure 6.** Relationship between the proportion of the total variance explained by the first eigenvalue and phenotypic integration in the original and simulated P-matrices.   

\pagebreak


## 7. Bootstrap   

### 7.1. Bootstrap reampling of the original data set.   

```{r}
# list of bootstraped data frames
# Boostrap samples were constrained to include at least seven 
# independent data, because positive-definite 6x6 P matrices 
# were needed for analysis.

B <- 1000 #number of bootstrap samples

list.boot <- function(x){
  pse.sal.list <- list(length(x))
  for(i in 1:length(x)) {
    pse.sal.list[[i]] <- x[[i]][c(sample(1:nrow(x[[i]]), size = 7, replace = FALSE),
                                  sample(1:nrow(x[[i]]), size = nrow(x[[i]])-7,
                                         replace = TRUE)),]
  }
  return(pse.sal.list)
}
megalist <- replicate(B, list.boot(x = sal.list), simplify = F)

# list of resampled species means
megabase <- lapply(megalist, FUN = function(x) do.call("rbind", x)) 
megapops <- lapply(megabase, FUN = function (x) {
  aggregate(x[, -3], by = list(x$pop), FUN = resum)[, -1]})
megameans <- lapply(megapops, FUN = function(x) {
  aggregate(x[, -2], by = list(x$sp), FUN = resum)[, -1]}) # Final base. 

# list of resampled P matrices
megaPpop <- lapply(megalist, FUN = function(x){
  lapply(x, FUN = function(y) {
    Z <- cov(y[, 10:15], use= "pairwise.complete.obs")
    return(Z)
  })
})

N <- numeric(21); for(i in 1:21) N[i] <- nrow(sal.list[[i]])

mergeP <- function(x){
    list(atro = x[[1]], calo = x[[2]], cinn = x[[3]],
         cusp = (x[[4]]*N[4]+x[[5]]*N[5])/(N[4]+N[5]), 
         eleg = (x[[6]]*N[6]+x[[7]]*N[7])/(N[6]+N[7]), 
         fulg = (x[[8]]*N[8]+x[[9]]*N[9])/(N[8]+N[9]), 
         guar = x[[10]], ioda = x[[11]], lava = x[[12]],
         long = x[[13]], mexi = x[[14]], mise = x[[15]],
         pall = x[[16]], pers = x[[17]], poly = x[[18]],
         purp = x[[19]], stac = x[[20]], thyr = x[[21]])
  }
       
megaPsp <- lapply(megaPpop, FUN = mergeP) # Final base. 
rm(megalist, megabase, megapops, megaPpop)

```
   
   

\pagebreak

### 7.2. Uncertainty around species means in the original mean trait space (morphospace).        
   
```{r}
library(ggplot2)

# projecting the resampled means into the phylogenetic
# principal component space
C <- vcv.phylo(Tsub1)[rownames(sp.means[, 8:13]), rownames(sp.means[, 8:13])]
temp <- phyl.vcv(as.matrix(sp.means[, 8:13]), C, 1)
a <- temp$alpha
A <- matrix(rep(a, 18), 18, 6, byrow = T)

PPC1.b <- lapply(megameans, FUN = function(x){
  (as.matrix(x[, 8:13]) - A) %*% (-1*phyPC$Evec[,1])
})
PPC2.b <- lapply(megameans, FUN = function(x){
  (as.matrix(x[, 8:13]) - A) %*% (-1*phyPC$Evec[,2])
})

new.ppc <- data.frame(PC1 = do.call("rbind", PPC1.b)*(-1),
                      PC2 = do.call("rbind", PPC2.b)*(-1))
DAT1 <- rbind(phyPC$S[, 1:2], new.ppc)
DAT1$sp <- rep(rownames(phyPC$S), B+1)
DAT1$synd <- rep(synd, B+1)

# plot
g1 <- ggplot(DAT1, aes(x = PC1, y = PC2, color = sp)) + geom_blank()
g1 <- g1 + scale_color_manual(values = colores) + stat_ellipse()
g1 <- g1 + geom_point(data = DAT1[1:18,]) 
g1 <- g1 + geom_text(data = DAT1[1:18, ], aes(x = PC1, y = PC2, label = sp),
                     hjust = 0, nudge_x = 0.08)
g1 <- g1 + guides(color = F) + theme_bw() + xlim(-2.5,2)
g1

```



**Figure 7.** Mean trait space (morphospace) of *Salvia* corolla traits, showing the first two phylogenetic principal components. Red dots indicate hummingbird-pollinated *Salvia* species, while blue dots correspond to bee-pollinated species. Ellipses indicate 95% confidence intervals, obtained through bootstrap resampling. The first four letters of spacies names are indicated.   

\pagebreak


### 7.3. Building a mean trait space (morphospace) in each bootstrap sample.   

```{r}
# Base of Phylogenetic Principal Components 
mega.phyPC <- lapply(megameans, FUN = function(x){
  rownames(x) <- x$sp
  new.phyPC <- phyl.pca(tree=Tsub1, Y = x[, 8:13])
  return(new.phyPC)
})

# Diferences in corolla morphology between pollination syndromes.
# PERMANOVA between groups
# Dispersion Test between groups
per.morpho <- lapply(mega.phyPC, FUN = function(x){
  x.dist <- dist(x$S)
  x.perm <- adonis2(x.dist ~ synd, permutations = 999)
  x.beta <- betadisper(x.dist, group=synd)
  x.beta2 <- permutest(x.beta, permutations = 999)
  return(c(F.permanova = x.perm[1, 4],
           P.permanova = x.perm[1, 5],
           F.betadisp = x.beta2$tab[1, 4],
           P.betadisp = x.beta2$tab[1, 6]))
})

per.morpho.dat <- as.data.frame(do.call("rbind", per.morpho))

# number of significant PERMANOVA results
sum(per.morpho.dat$P.permanova < 0.05)

# number of significant dispersion test results
sum(per.morpho.dat$P.betadisp < 0.05)

layout(matrix(1:4, 2, 2))
hist(per.morpho.dat$F.permanova, main = NULL, 
     xlab = "F permanova", col = "grey")
hist(per.morpho.dat$P.permanova, main = NULL, 
     xlab = "P permanova", col = "grey")
hist(per.morpho.dat$F.betadisp, main = NULL, 
     xlab = "F dispersion", col = "grey")
hist(per.morpho.dat$P.betadisp, main = NULL, 
     xlab = "P dispersion", col = "grey")
layout(1)

```
   
   
   
**Figure 8.** PERMANOVA and Dispersion Test statistics of differences in mean traits between pollination syndromes. Histograms are basen on 1000 bootstrap samples of the original data.

\pagebreak

### 7.4. Uncertainty around P matrices in the original variance-covariance space.       
   
```{r fig.height = 8}
library(gridExtra)

## reprojecting bootstraped P matrices into the original covariance space
mega.proj <- lapply(megaPsp, FUN = function(x){
  unc.pmat <- c(salvia_p, x)
  names(unc.pmat) <- NULL
  dist.unc.pmat <- as.dist(MatrixDistance(cov.x = unc.pmat, distance = "RiemannDist"))
  sup.unc <- as.matrix(dist.unc.pmat)[19:36,-c(19:36)]
  proj.unc <- mmds.project(mmds=biommds, sup = sup.unc, pc = 17)
  proj.unc
})

mega.proj.dat <- do.call("rbind", lapply(mega.proj, FUN = function(x){
                           x$coord[, 1:3]}))
mega.proj.dat$sp <- rep(names(salvia_p), B)
DAT2 <- rbind(biommds$coord[, 1:3], mega.proj.dat[,1:3])
DAT2$sp <- c(rownames(biommds$coord), mega.proj.dat$sp)
DAT2$synd <- rep(synd, B+1)

# plot confidence ellipses
g2 <- ggplot(DAT2, aes(x = PC1, y = PC2, color = sp)) + geom_blank()
g2 <- g2 + scale_color_manual(values = colores) + stat_ellipse()
g2 <- g2 + geom_point(data = DAT2[1:18,]) 
g2 <- g2 + geom_text(data = DAT2[1:18, ], aes(x = PC1, y = PC2, label = sp),
                     hjust = 0, nudge_x = 0.2)
g2 <- g2 + guides(color = F) + theme_bw() + xlab("PCo1") + ylab("PCo2")

g3 <- ggplot(DAT2, aes(x = PC1, y = PC3, color = sp)) + geom_blank()
g3 <- g3 + scale_color_manual(values = colores) + stat_ellipse()
g3 <- g3 + geom_point(data = DAT2[1:18,]) 
g3 <- g3 + geom_text(data = DAT2[1:18, ], aes(x = PC1, y = PC3, label = sp),
                     hjust = 0, nudge_x = 0.2)
g3 <- g3 + guides(color = F) + theme_bw() + xlab("PCo1") + ylab("PCo3")

grid.arrange(g2, g3)
   
```
   
   
   
**Figure 9.** Variance-covariance space of *Salvia* **P**-matrices, showing the first three principal coordinates. Red dots indicate hummingbird-pollinated *Salvia* species, while blue dots correspond to bee-pollinated species. Ellipses indicate 95% confidence intervals, obtained through bootstrap resampling. The first four letters of spacies names are indicated.

\pagebreak

### 7.5. Building a variance-covariance space in each bootstrap sample.    

```{r}
# Riemanian distances
mega.rdist <- lapply(megaPsp, FUN = function(x){
  as.dist(MatrixDistance(cov.x = x, distance = "RiemannDist")) 
})

# Principal Coordinate Analysis (PCoA)
mega.pcoa <- lapply(mega.rdist, FUN = function(x){
  mmds(active = as.matrix(x), pc = 17)
})

# PERMANOVA & Dispersion
per.covar <- lapply(mega.rdist, FUN = function(x){
  x.perm <- adonis2(x ~ synd, permutations = 999)
  x.beta <- betadisper(x, group=synd)
  x.beta2 <- permutest(x.beta, permutations = 999)
  return(c(F.permanova = x.perm[1, 4],
           P.permanova = x.perm[1, 5],
           F.betadisp = x.beta2$tab[1, 4],
           P.betadisp = x.beta2$tab[1, 6]))
})

per.covar.dat <- as.data.frame(do.call("rbind", per.covar))

# number of significant PERMANOVA results
sum(per.covar.dat$P.permanova < 0.05)

# number of significant dispersion test results
sum(per.covar.dat$P.betadisp < 0.05)

layout(matrix(1:4, 2, 2))
hist(per.covar.dat$F.permanova, main = NULL, 
     xlab = "F permanova", col = "grey")
hist(per.covar.dat$P.permanova, main = NULL, 
     xlab = "P permanova", col = "grey")
abline(v = 0.05, col = "red")
hist(per.covar.dat$F.betadisp, main = NULL, 
     xlab = "F dispersion", col = "grey")
hist(per.covar.dat$P.betadisp, main = NULL, 
     xlab = "P dispersion", col = "grey")
abline(v = 0.05, col = "red")
layout(1)

```



**Figure 10.** PERMANOVA and Dispersion Test statistics of differences in covariance structure between pollination syndromes. Histograms are basen on 1000 bootstrap samples of the original data. Red lines indicate P = 0.05.

\pagebreak

### 7.6. Comparison of morphological and variance-covariance spaces in bootstrap samples.   

```{r}
dist.coph <- cophenetic.phylo(Tsub1)
mega.compspaces <- list(length = B)
for(i in 1:B){
  obs.value <- partial.cor(xdis = as.matrix(mega.rdist[[i]]), 
                           ydis = as.matrix(dist(mega.phyPC[[i]]$S, upper = T, diag = T)), 
                           zdis = dist.coph)
  partial.mantel <- replicate(B, 
                              partial.cor(xdis = as.matrix(mega.rdist[[i]])[
                                phyloPermute(phy=Tsub1), phyloPermute(phy=Tsub1)],
                                ydis = as.matrix(dist(mega.phyPC[[i]]$S, upper = T, diag = T)),
                                zdis = dist.coph))
  p.value <- length(partial.mantel[partial.mantel>obs.value])/B
  mega.compspaces[[i]] <- c(obs.value, p.value)
}
mega.compspaces <- do.call("rbind", mega.compspaces)

# 95% confidence interval for partial Mantel correlation (r)
quantile(mega.compspaces[, 1], probs = c(0.025, 0.975))

# number of significant partial Mantel test results
sum(mega.compspaces[, 2]<0.05) 

# plot
layout(matrix(1:2, 1, 2))
hist(mega.compspaces[, 1], main = NULL, xlab = "r", col = "grey")
hist(mega.compspaces[, 2], main = NULL, xlab = "P", col = "grey")
abline(v = 0.05, col = "red")
layout(1)

```



**Figure 11.** Partial Mantel test with phylogenetic permutation for the comparison between mean trait space and variance-covariance space. Left: partial correlation value (r) between spaces in 1000 bootstrap samples. Right: Significance after 100 permutations"

\pagebreak

### 7.7. Numerical simulations in each bootstrap sample.   

#### 7.7.1. Average vector correlation between principal coordinates and proportionality changes.   

```{r}
# Creating new matrices from the original P-matrices.
matrix.index <- rep(names(salvia_p), 2*B)

mega.prop.mat <- lapply(megaPsp, FUN = function(x){
  Z <- list(36)
  for(i in 1:36){
    Z[[i]] <- prop_matrix(x[[matrix.index[i]]], pc = prop.values[i])
    names(Z)[i] <- paste(matrix.index[i], prop.values[i], sep = "_")
  }
  Pmat.add1 <- c(x, Z) 
  distances1 <- as.dist(MatrixDistance(cov.x = Pmat.add1, distance = "RiemannDist"))
  SUP1 <- as.matrix(distances1)[19:54,-c(19:54)]
  pru1 <- mmds.project(mmds=biommds, sup = SUP1, pc = 17)
  return(pru1)
})

# vector correlation. Proportionality simulations
angles.prop <- list(length = B)
for(i in 1:B){
  ori1 <- paste(names(megaPsp[[1]]), 0.5, sep = "_")
  fin1 <- paste(names(megaPsp[[1]]), 2, sep = "_")
  ANG1 <- list(18)
  for(j in 1:18){ 
    ANG1[[j]] <- angle.calc(from = ori1[j], to=fin1[j], 
                            coord = mega.prop.mat[[i]]$coord, 
                            dim = 17)
  }
  ANG.PROP <- do.call(rbind, ANG1)
  vector.corr1 <- colMeans(ANG.PROP)[1:7]
  vector.angles <- colMeans(ANG.PROP)[18:24]
  angles.prop[[i]] <- c(vector.corr1, vector.angles)
}
angles.prop <- do.call ("rbind", angles.prop)

# Which principal coordinate has higher absolute correlation with the direction of 
# proportionality changes?
which.pco.prop <- numeric(B)
for(i in 1:B){
which.pco.prop[i] <- as.character(which.max(abs(angles.prop[i, 1:7])))
}
table(which.pco.prop)

# plot
dat.angle1 <- data.frame(PCo = sort(rep(paste("PCo", c(1:7), sep = "."), B)),
                         corr = c(angles.prop[, 1], angles.prop[, 2], angles.prop[, 3],
                                  angles.prop[, 4], angles.prop[, 5], angles.prop[, 6],
                                  angles.prop[, 7]))
plot(corr~PCo, data = dat.angle1, ylab = "angle correlation", 
     col = "grey")

```



**Figure 12.** Correlations between the average vector of proportionality changes and the first seven principal coordinate axes of the variance-covariance space. Vector correlations were obtained in 1000 bootstrap samples.

\pagebreak

#### 7.7.2. Average vector correlation between principal coordinates and integration changes.   

```{r}
# Creating new matrices from the original P-matrices.
mega.int.mat <- lapply(megaPsp, FUN = function(x){
  Z <- list(36)
  for(i in 1:36){
    Z[[i]] <- bs_matrix(x[[matrix.index[i]]], prop.var = leading.ei[i])
    names(Z)[i] <- paste(matrix.index[i], leading.ei[i], sep = "_")
  }
  Pmat.add2 <- c(x, Z) 
  distances2 <- as.dist(MatrixDistance(cov.x = Pmat.add2, distance = "RiemannDist"))
  SUP2 <- as.matrix(distances2)[19:54,-c(19:54)]
  pru2 <- mmds.project(mmds=biommds, sup = SUP2, pc = 17)
  return(pru2)
})

# vector correlation. Integration simulations.
angles.int <- list(length = B)
for(i in 1:B){
  ori1 <- paste(names(megaPsp[[1]]), 0.25, sep = "_")
  fin1 <- paste(names(megaPsp[[1]]), 0.85, sep = "_")
  ANG2 <- list(18)
  for(j in 1:18){ 
    ANG2[[j]] <- angle.calc(from = ori1[j], to=fin1[j], 
                            coord = mega.int.mat[[i]]$coord, 
                            dim = 17)
  }
  ANG.INT <- do.call(rbind, ANG2)
  vector.corr1 <- colMeans(ANG.INT)[1:7]
  vector.angles <- colMeans(ANG.INT)[18:24]
  angles.int[[i]] <- c(vector.corr1, vector.angles)
}
angles.int <- do.call("rbind", angles.int)

# Which principal coordinate has higher correlation with the direction of 
# integration changes?
which.pco.int <- numeric(B)
for(i in 1:B){
which.pco.int[i] <- as.character(which.max(abs(angles.int[i, 1:7])))
}
table(which.pco.int)

# plot
dat.angle2 <- data.frame(PCo = sort(rep(paste("PCo", c(1:7), sep = "."), B)),
                         corr = c(angles.int[, 1], angles.int[, 2], angles.int[, 3],
                                  angles.int[, 4], angles.int[, 5], angles.int[, 6],
                                  angles.int[, 7]))
plot(corr~PCo, data = dat.angle2, ylab = "angle correlation", 
     col = "grey")

```



**Figure 13.** Correlations between the average vector of integration changes and the first seven principal coordinate axes of the variance-covariance space. Vector correlations were obtained in 1000 bootstrap samples.


